Continuous Deployment and Schema Evolution \\in SQL Databases

### 2015-01-21 paper.tex
Improved draft, added 'approach' and 'implementation status' sections.

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

@@

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
% Read the url.sty source comments for usage information. Basically,
% \url{my_url_here}.

@@

\usepackage{color}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{url}

@@

SQL databases are an integral part of many (web) applications. They provide strong guarantees about the storage and retrieval of data which makes them useful for various software applications. However the environment of software engineers and release engineers is drastically changing: shorter development iterations, more frequently deployed to production, or even automatically deployed to production several times a day. For release engineers this brings additional challenges. In particular when evolving the schema (tables, columns, etc) of the SQL database.

This can essentially be distilled into two separate problems:
\begin{itemize}
  \item{Given the deployment of a new version of a web application, depending on the deployment strategy it might be required that both the old and the new version of the application run in parallel (\textit{Rolling Upgrades}, or \textit{Canary Releases}). This means that the schema on the SQL database needs to be compatible with the schema expected by both the old and the new application.}
  \item{Operations which modify the schema in SQL databases (\textit{DDL statements}), depending on the SQL database implementation may be either non-blocking, or blocking in nature. In case of non-blocking behaviour, queries issued by the application for instance may proceed, whereas with blocking behaviour these queries may be put on hold for the duration of the schema modification. This can range anywhere from milliseconds to hours depending on the operation, the size of the affected table, and database implementation.}
\end{itemize}

These two problems make releasing often, problematic for applications with no service windows. To address this problem we aim to implement our own proof-of-concept tooling, which would allow its users to not only version their database schema, but also evolve their database schema without affecting older clients because of changes in the database schema or using blocking schema operations.

@@

Before we decided to develop our own tooling, we examined existing approaches, tools and research which aims to solve these problems. A good starting point is the seminal book ``Continuous Delivery''~\cite{Humble:2010:CDR:1869904}. This book describes three high-level approaches to deal with migrating data without requiring downtime:

\begin{itemize}
  \item{\textbf{Approach 1}: Copy the database and start recording transactions. When the copying has completed, apply the schema changes. Finally replay all the recorded database transactions to reach an equivalent database.}
  \item{\textbf{Approach 2}: When using Blue-Green deployments, one possible approach is to copy the database to the alternative production environment and apply the schema operations there. Meanwhile these two databases need to be kept in sync, or the original must be in a read-only state.}
  \item{\textbf{Approach 3}: Another approach is performing schema operations separately from your application deployments using operations which are non-blocking. As a consequence the application must be able to deal with either the old or the new database schema being in use, and formulate queries accordingly.}
\end{itemize}

@@

\subsubsection{Openark Kit, Percona Toolkit}
The Openark Kit~\cite{OpenarkKit} and Percona Toolkit~\cite{PerconaToolkit} are tools designed to help DBAs perform administrative tasks on MySQL databases. \textit{oak-online-alter-table} and \textit{pt-online-schema-change} respectively are commands which follow a variation of approach 1. They create an empty copy of the a specified table and then apply schema operations on the new table. These commands then proceed to install triggers on both tables, which insert, update, or delete the corresponding record in the other table when a record in their own table is inserted, updated, or deleted. A background task then iterates over all rows in the original table (in batches) and lazily copies rows from the original table to the new table. When all rows have been copied the original table and new table switch names, effectively ensuring that clients now use the new table when querying the old table name.

\subsubsection{TableMigrator}
TableMigrator~\cite{TableMigrator} is a tool which works very similar to Openark Kit and Percona Toolkit. But instead of using triggers to synchronize the two tables with each other. TableMigrator requires tables to have an additional column which stores the date and time that record was last modified. Using this information corresponding rows in both tables are synchronized with each other by passing over all rows in multiple passes. Once a pass can be done relatively quickly, TableMigrator acquires an exclusive write lock --- blocking the entire table --- and proceeds to do a final pass. Once this has been done, it atomically swaps both table names before releasing the write lock.

\subsubsection{Online Schema Change \& Large Hadron Migrator}
Facebook's Online Schema Change (\textit{OSC})~\cite{FacebookOSC} and Soundcloud's Large Hadron Migrator (\textit{LHM})~\cite{SoundcloudLHM} are both tools which closely follow approach 1. These tools create a full copy of a table (including its contents) and record modifications to rows in the original table using triggers into a special \textit{deltas} table. Once they have copied the original table, they apply the specified schema operations to the new table before proceeding to replay the recorded modifications from the \textit{deltas} table onto the new table. When the \textit{deltas} table is almost empty, these tools acquire an exclusive write lock, apply the last recorded modifications and then atomically switch the names of both tables.

@@

\subsubsection{PRISM?}
\textbf{Should I cover PRISM and SMOs here?}

\subsubsection{Imago}
Imago is a tool which follows approach 2. With Imago there are two production environments --- called "universes" by the authors. Imago integrates at both ingress point (the load balancer) to be able to switch users from one universe to the other, and at the egress point where the (web) application sends queries to the SQL database. When a new version of the application (requiring a schema change) needs to be deployed, Imago bootstraps the inactive production environment with a copy of the database, applies the required schema operation and then copies over the data from the old universe to the new universe. When this process is almost complete, it enforces a short period of read-only access to the SQL database while it is switching users from the old environment to the new using the load balancer.

@@

\begin{itemize}
  \item{None of the tools used in practice support foreign keys and primarily focus on MySQL databases. Imago makes no claims with regards to support for foreign keys, but does impose restrictions on how to deploy applications, and makes assumptions on the read/write ratio of database queries.}
  \item{Tools which do not handle the deployment of the application themselves, also force the developers of the application to ensure that their application is able to deal with the schema being in either the old or the new state. All tools atomically rename the modified table, but the application itself is left to deal with modifying queries to accommodate added or removed columns, constraints, or tables at the time of the atomic table rename operation.}
\end{itemize}

@@

We feel that the current situation is an impediment to adoption and growth of \textit{Continuous Deployment}. Although \textit{Continuous Deployment} has made great improvements on workflows, team dynamics, and tooling for deploying and monitoring applications, the area of schema evolution in databases has received little attention. Therefore we aim to address these challenges by developing our own tool which adheres to the following principles:

\begin{itemize}
  \item{Do not restrict release engineers, or tell them how to deploy their applications. Release engineers have their own tools to perform deployments.}
  \item{If an application requires a SQL database, it should be exposed as one or more database schemas simultaneously which are synchronized with each other in a transactionally safe manner. Each client should connect to the database using one particular schema. This allows for \textit{Canary/Feature Deployments}, as well as \textit{Blue-Green deployments} and \textit{Rolling upgrades}.}
  \item{Versioning and rolling back are important for release engineers. Any tool which modifies the database schema should allow for versioning and be able to rollback to previous versions.}
  \item{Foreign keys are an integral part of SQL databases and should be supported.}
\end{itemize}

\subsection{Approach}

@@

\begin{itemize}
  \item{Current state of implementation}
  \item{Road ahead?}
\end{itemize}

@@

\begin{itemize}
  \item{Experiments at Magnet.me}
  \item{Figures}
\end{itemize}

>>>>

% *** CITATION PACKAGES ***
%
\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

@@

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
% Read the url.sty source comments for usage information. Basically,
% \url{my_url_here}.

@@

@@

SQL databases are an integral part of many (web) applications. They provide strong guarantees about the storage and retrieval of data which makes them useful for various software applications. However the environment of software engineers and release engineers is drastically changing: shorter development iterations, more frequently deployed to production, or even automatically deployed to production several times a day. For release engineers this brings additional challenges. In particular when evolving the schema (tables, columns, foreign keys, etc) of the SQL database either forwards or backwards without loss of data.

This can essentially be distilled into two separate problems:
\begin{itemize}
  \item{Given the deployment of a new version of a web application, depending on the deployment strategy it might be required that both the old and the new version of the application run in parallel (\textit{Rolling Upgrades}, or \textit{Canary Releases}). This means that the schema on the SQL database needs to be compatible with the schema expected by both the old and the new application.}
  \item{Operations which modify the schema in SQL databases (\textit{DDL statements}), depending on the SQL database implementation may be either non-blocking, or blocking in nature. In case of non-blocking DDL statements queries issued by the application for instance may proceed, whereas with blocking DDL statements these queries may be put on hold for the duration of the schema modification. This disruption can range anywhere from milliseconds to hours depending on the DDL statement, the size of the affected table, and database implementation. One such example would be adding a new non-nullable column to a PostgreSQL database. This operation is blocking in nature --- preventing other queries from running while it is being executed --- and when applied to a table of 100 million rows (in our tests) blocked other queries from running on that table for 5 minutes.}
\end{itemize}

When service windows are unavailable, these two problems make releasing applications often, problematic for release engineers. To address this problem we aim to find or implement tooling which would allow its users to not only version their database schema, but also evolve their database schema without affecting older and still running applications.

@@

Before we decided to develop tooling to solve this ourselves, we examined existing approaches, tools, and research which aims to solve these problems. A good starting point is the seminal book ``Continuous Delivery''~\cite{Humble:2010:CDR:1869904}. This book describes three high-level approaches to deal with migrating data without data loss or requiring downtime:

\begin{itemize}
  \item{\textbf{Approach 1}: Copy the database or schema and start recording transactions. When the copying has completed, apply the schema changes to the copied database or schema. Finally replay all the recorded database transactions on the copied database or schema to reach an equivalent state. Once this state has been achieved, it must either be maintained by continuing to record and playback transaction, or one must make the switch from the old to the new database or schema.}
  \item{\textbf{Approach 2}: When using Blue-Green deployments, two instances of the database are available. In order to evolve the schema one must apply the schema operations to the inactive database. In the mean time these two databases need to be kept synchronized. This can be done by using approach 1, or somehow ensure that the table under change remains in a read-only state until the active and inactive databases have switched roles.}
  \item{\textbf{Approach 3}: The last is an approach where schema operations are performed separately from your application deployments using operations which are non-blocking. As a consequence the application must be able to deal with either the old or the new database schema being in use, and formulate queries accordingly.}
\end{itemize}

@@

\subsubsection{Percona Toolkit~\cite{PerconaToolkit} \& Openark Kit~\cite{OpenarkKit}}
are tools designed to help DBAs perform administrative tasks on MySQL databases. \textit{oak-online-alter-table} and \textit{pt-online-schema-change} respectively are commands which follow a variation of approach 1. They create an empty copy of the a specified table and then apply schema operations on the new table. These commands then proceed to install triggers on both tables, which insert, update, or delete the corresponding record in the other table when a record in their own table is inserted, updated, or deleted. A background task then iterates over all rows in the original table (in batches) and lazily copies rows from the original table to the new table. When all rows have been copied the original table and new table switch names, effectively ensuring that clients now use the new table when querying the old table name.

\subsubsection{TableMigrator~\cite{TableMigrator}}
is a tool which works very similar to Openark Kit and Percona Toolkit. But instead of using triggers to synchronize the two tables with each other. TableMigrator requires tables to have an additional column which stores the date and time that record was last modified. Using this information corresponding rows in both tables are synchronized with each other by passing over all rows in multiple passes. Once a pass can be done relatively quickly, TableMigrator acquires an exclusive write lock --- blocking the entire table --- and proceeds to do a final pass. Once this has been done, it atomically swaps both table names before releasing the write lock.

\subsubsection{SoundCloud's Large Hadron Migrator~\cite{SoundcloudLHM} \& Facebook's Online Schema Change~\cite{FacebookOSC}}
are both tools which closely follow approach 1. These tools create a full copy of a table (including its contents) and record modifications to rows in the original table using triggers into a special \textit{deltas} table. Once they have copied the original table, they apply the specified schema operations to the new table before proceeding to replay the recorded modifications from the \textit{deltas} table onto the new table. When the \textit{deltas} table is almost empty, these tools acquire an exclusive write lock, apply the last recorded modifications and then atomically switch the names of both tables.

@@

% Technically we could go without this subsubsection...
\subsubsection{PRISM}
PRISM~\cite{Curino:2010:URI:1921071.1921078, Curino:2009:ADS:1656437.1656444} is a tool which is a part of a larger framework. Although PRISM does not solve the problem of downtime when performing operations on the database schema, it does aim to shorten it. PRISM is designed to reason about the database schema changes, and infer how to rewrite existing queries to the new resulting database schema. In addition it can generate the migration and rollback scripts which will transform the current database schema to the new database schema and back. The most influential concept of PRISM is the definition of several new operations to evolve a database schema. These described Schema Modification Operations (\textit{SMOs}), offer ways to split or merge tables into multiple tables (structurally or content-wise).

\subsubsection{Imago}
Imago~\cite{Dumitras:2009:WUF:1656980.1657005} is a tool which follows approach 2. With Imago there are two production environments --- called "universes" by the authors. Imago integrates at both ingress point (the load balancer) to be able to switch users from one universe to the other, and at the egress point where the (web) application sends queries to the SQL database. When a new version of the application (requiring a schema change) needs to be deployed, Imago bootstraps the inactive production environment with a copy of the database, applies the required schema operation and then copies over the data from the old universe to the new universe. When this process is almost complete, it enforces a short period of read-only access to the SQL database while it is switching users from the old environment to the new, using the load balancer. Imago makes no claims in regards to its support for foreign keys.

@@

\begin{itemize}
  \item{None of the tools used in practice support foreign keys and primarily focus on MySQL databases.}
  \item{Tools which do not handle the deployment of the application themselves, all force the developers of the application to ensure that their application is able to deal with the database schema being in either the old or the new state. All tools atomically rename the modified table, but the software developers of the application are left to deal with modifying queries to accommodate added or removed columns, constraints, or tables at the time of the atomic table rename operation.}
\end{itemize}

@@

We feel that the current situation is an impediment to adoption and growth of \textit{Continuous Deployment}. Although \textit{Continuous Deployment} has made great improvements on workflows, team dynamics, and tooling for deploying and monitoring applications, the area of schema evolution in databases has received little attention and no suitable solution currently exists. Therefore we aim to address these challenges by developing a new tool which adheres to the following restrictions:

\begin{itemize}
  \item{A solution should not put any restrictions on how to deploy the application, or which methods of deployment that can be used: \textit{Blue-Green deployments}, \textit{Rolling upgrades}, \textit{Canary releases}, etc.}
  \item{The contents of the database must be accessible through one or more database schemas, which are continuously kept synchronized with each other.}
  \item{Clients connected to the database must be able to ``view'' the contents of the database according to an active database schema of their choice. Their ``view'' should only contain tables, columns, foreign keys, etc which are present in their chosen database schema. Even when more than one database schema is active at the same time.}
  \item{Versioning and rolling back are important for release engineers. A solution should support versioning the changes being made to the database schema, and be able to rollback to a previous version.}
  \item{Foreign keys are an integral part of SQL databases and should be supported.}
\end{itemize}

\subsection{Solution direction}

We have come up with an approach which combines approaches 1 and 3. The previously discussed tools from practice all use approach 1, and prove that this is a viable way for performing simple evolutions of the database schema. These tools all create a single ghost table, whereas we aim to create one ghost table for each table that is under change, and ghost tables for all tables which in turn are dependent on these tables under change. This approach ensures that we do not violate any previously defined foreign keys.

\textbf{Show image an example here?}

In addition, the tools from practice all allow release engineers to switch the table names manually after the the original and the ghost table have been synchronized. This is good since it allows for two or more versions of the application to co-exist on the same database as long as each version correctly queries the correct (ghost)table. Although the current tools from practice do not solve this specific problem, we think that this is easily solved using a small wrapper around database drivers for most programming languages. Alternatively, for software engineers working on the application we could expose which (ghost)table should be queried for a specific version of the database schema, when writing SQL queries.

@@

\subsection{Current status}

We are implementing a new tool --- QuantumDB~\cite{QuantumDB} --- which aims to address the need for non-blocking schema evolution, and remove obstacles for teams who want to adopt \textit{Continuous Deployment} but struggle because of their SQL database(s). Although not yet usable in production environments, we have implemented enough features at the moment to take a step back and evaluate QuantumDB in an small experiment.

QuantumDB currently consists of two parts: 
\begin{itemize}
  \item{\textbf{core} is the program responsible for performing the actual migration tasks on the database. It does this by creating ghost tables of tables under change, using triggers to keep the ghost tables and the original tables synchronized with each other, and a background task which copies the data from the original tables to the ghost tables in small batches to prevent overloading, and degrading the performance of the database.}
  \item{\textbf{driver} is a Java JDBC driver which can wrap another JDBC driver. When used, the application developers can specify a schema version on which their application operates. The driver then rewrites SQL queries sent to it by the application, and passes them on to the wrapped database JDBC driver which in turn sends it to the database. When rewriting queries, it replaces table names with the names of the associated (ghost)table names if applicable.}
\end{itemize}

With QuantumDB we can currently take one of the active database schemas, apply an arbitrary number of schema operations to it, and expose the resulting database schema in addition to the original database schemas. Once none of the applications are still connected to a particular version of the database schema, it can safely be discarded in a background task. Effectively this can be described as A/B testing the database schema.

\subsection{Road ahead}

As stated before, the implementation of QuantumDB is not yet ready for production. To be able to use QuantumDB effectively in production environments, we still need to implement some features.

\begin{itemize}
  \item{QuantumDB can already deal with foreign keys referring from tables under change to other tables not under change. To fully support foreign keys QuantumDB would also need to be able to deal with foreign keys referring to tables under change.}
  \item{QuantumDB currently only supports PostgreSQL, but should prove easily extensible to other popular SQL databases since most of these have support for triggers and procedures.}
  \item{The driver currently uses a very rudimentary query rewriting algorithm to replace table names under change with the names of ghost tables. It does not yet support the entire PostgreSQL query syntax.}
  % Technically speaking this next one is optional.%
  \item{Adding support for SMOs. These operations allow for more complex schema operations on tables, and should enable software engineers to make more elaborate changes to the database schema.}
  \item{Evaluate QuantumDB more extensively in use cases at different organizations and in production environments.}
\end{itemize}

@@

\begin{itemize}
  \item{Experiment setup}
  \item{Results}
\end{itemize}
